Research Experience

### 2010-03-24 doc/thesis/thesis.tex
need to justify factoring importance density

%\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}

@@

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise and noise in $x$ can drive
changes in $y$, it is clear that $Pr(x(t) | u[0:T]$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a proposal distribution, and $N_p$
particles need to be drawn from that distribution, $\alpha(x)$:
\begin{equation}
\{[\hat{x}_i(0),w_i] : \hat{x}_i(0) \sim \alpha(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

@@

Note that all the probability's implicitly depend on $u[0:T]$, so those terms 
will be left off for simplicity.
Once the probability, $Pr(\hat{x}(T) | \hat{x}[0:T-1], y[0:T-1]$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between before a measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_x$ should increase uncertainty as
time without a measurement passes.

@@

\begin{equation}
w_i(T) \propto \frac{Pr(x_i[0:T] | y[0:T])}{q(x_i)}
\end{equation}

where $q$ is called an \emph{importance density}, which means it decides where
the support points for $\hat{x}(T)$ are located. To remove the bias due to the
location of the support points, we divide by $q(\hat{x}_i)$. By dividing by 
the density of the support points (particles), the location of the points are removed
from the probability density. As a result the weight is dependent solely based
on $Pr(x_i[0:T] | y[0:T]$, the probability of the $i^{th}$ particle's measurements
being different from $y[0:T]$ due to noise alone.
An example of an importance density would be drawing a large
number of points from the standard normal, $N(0,1)$ and then weighting each point, $l$ by 
$1/\beta(l), \beta \sim N(0,1)$. Of course if there is a far off peak in
the posterior that $q$ does not allocate support points in, there will essentially
be a quantization error, and that part of the density can't be modeled.

For this work we used $q(x_i) = Pr(x_i(T) | x_i[0:T-1]) = Pr(x_i(T) | x_i(T-1))$,
based on the Markov assumption, and the belief that the state space model is 
at least close to correct.  
%<stopped here>
This has the added benefit that it is not necessary to
re-draw points from a $q$ function at every time step, which is faster. 
this the definition of the weight, so when weights are updated it is necessary
to divide by the proposal distributionto converge
to the true distribution it is necessary to divide by the proposal distribution.

>>>>

%\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}

@@

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise and noise in $x$ can drive
changes in $y$, it is clear that $Pr(x(t) | u[0:T])$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a proposal distribution, and $N_p$
particles need to be drawn from that distribution, $\alpha(x)$:
\begin{equation}
\{[\hat{x}_i(0),w_i] : \hat{x}_i(0) \sim \alpha(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

@@

Note that all the probability's implicitly depend on $u[0:T]$, so those terms 
will be left off for simplicity.
Once the probability, $Pr(\hat{x}(T) | \hat{x}[0:T-1], y[0:T-1])$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between before a measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_x$ should increase uncertainty as
time without a measurement passes.

@@

\begin{equation}
\label{weightfunc}
w_i(T) \propto \frac{Pr(x_i[0:T] | y[0:T])}{q(x_i)}
\end{equation}

where $q$ is called an \emph{importance density}, meaning it decides where
the support points for $\hat{x}(T)$ are located. To remove the bias due to the
location of the support points, we divide by $q(\hat{x}_i)$. By dividing by 
the density of the support points (particles), the effect of the particle 
distribution may be removed from the posterior density. As a result the weight
is dependent solely based
on $Pr(x_i[0:T] | y[0:T])$, the probability of the $i^{th}$ particle's measurements
being different from $y[0:T]$ due to noise alone.
An example of an importance density would be drawing a large
number of points from the standard normal, $N(0,1)$ and then weighting each point, $l$ by 
$1/\beta(l), \beta \sim N(0,1)$. Of course if there is a far off peak in
the posterior that $q$ does not allocate support points in, there will essentially
be a quantization error, and that part of the density can't be modeled.

For this work we used $q(x_i) = Pr(x_i(T) | x_i[0:T-1]) = Pr(x_i(T) | x_i(T-1))$,
based on the Markov assumption, and the belief that the state space model is 
able to approximate the true state. The weight is also based on $Pr(x_i[0:K] | y[0:K])$,
which may be broken up as follows.
\begin{eqnarray}
Pr(x[0:T] | y[0:T]) & = & \frac{Pr(y[0:T], x[0:T])}{Pr(y[0:T])} \nonumber \\
 & = & \frac{Pr(y(T), x[0:T] | y[0:T-1]) \cancel{Pr(y[0:T-1])}}{Pr(y(T) | y[0:T-1]) \cancel{Pr(y[0:T-1])}} \nonumber \\
 & = & \frac{Pr(y(T)| x[0:T], y[0:T-1]) Pr(x[0:T] | y[0:T-1])}{Pr(y(T) | y[0:T-1]) } \nonumber \\
 & = & \frac{Pr(y(T)| x[0:T], y[0:T-1]) Pr(x(T) | x[0:T-1], y[0:T-1]) Pr(x[0:T-1] | y[0:T-1])}{Pr(y(T) | y[0:T-1])} \nonumber \\
\end{eqnarray}

Using the assumption that $y(t)$ is fully constrained by $x(t)$,
and that $x(t)$ is fully constrained by $x(t-1)$, we are able to
make the reasonably good assumptions that:
\begin{equation}
Pr(y(T) | x[0:T], y[0:T-1]) = Pr(y(T) | x(T))
\end{equation}

\begin{equation}
Pr(x(T) | x[0:T], y[0:T-1]) = Pr(x(T) | x(T-1))
\end{equation}

Additionally, for the particle filter $y(T)$ and $y[0:T-1]$ are 
given, and therefore constant across all particles. Thus $Pr(x[0:T] | y[0:T]$
may be simplified to:
\begin{eqnarray}
Pr(x[0:T] | y[0:T]) & = & \frac{Pr(y(T)| x[0:T], y[0:T-1]) Pr(x(T) | x[0:T-1], y[0:T-1]) 
            Pr(x[0:T-1] | y[0:T-1])}{Pr(y(T) | y[0:T-1])} \nonumber \\
& = & \frac{Pr(y(T)| x(T)) Pr(x(T) | x(T-1)) Pr(x[0:T-1] | y[0:T-1])}{Pr(y(T) | y[0:T-1])} \nonumber \\
& \propto & Pr(y(T)| x(T)) Pr(x(T) | x(T-1)) Pr(x[0:T-1] | y[0:T-1])
\end{eqnarray}

Plugging this into \ref{weightfunc} leads to:
\begin{eqnarray}
w_i(T) & \propto & \frac{Pr(y(T)| x(T)) Pr(x(T) | x(T-1)) Pr(x[0:T-1] | y[0:T-1])}{q(x(T) | y[0:T])q(x[0:T-1] | y[0:Tjk:wq
} \nonumber \\
& \propto & Pr(y(T)| x(T)) Pr(x[0:T-1] | y[0:T-1]) \\
& \propto & \frac{Pr(y(T)| x(T)) \cancel{Pr(x(T) | x(T-1))} Pr(x[0:T-1] | y[0:T-1])}{\cancel{Pr(x(T) | x(T-1))}} \nonumber \\
& \propto & Pr(y(T)| x(T)) Pr(x[0:T-1] | y[0:T-1]) \\
\end{eqnarray}

and of course, since ....wait?
%<stopped here>
This has the added benefit that it is not necessary to
re-draw points from a $q$ function at every time step, which is faster. 
this the definition of the weight, so when weights are updated it is necessary
to divide by the proposal distributionto converge
to the true distribution it is necessary to divide by the proposal distribution.
